Weak interspecific interactions in a sagebrush steppe: Evidence from observational data, population models, and a removal experiment

### 2016-09-20 removals/manuscript/removal_ms.tex
Manuscript edits

To quantify the short-term population-level consequences of removals, we used an IBM to project the cover of each species in each quadrat one year ahead using versions of the model with and without removal treatment effects. Our models tended to under-predict population growth (Fig. \ref{fig:CoverChange}) and population size (Fig. \ref{fig:IBM1step}) for all species, but the difference between predictions from the baseline and removal models varied among species. For \textit{A. tripartita}, both observed and predicted population growth rates were close to zero, and differences in model predictions were small. \textit{H. comata} experienced greater variation in growth rates, but the inclusion of removal treatment effects had little effect on population growth rates. For \textit{Poa secunda} and especially \textit{P. spicata}, the model with removal effects consistently predicted higher population growth than the baseline model (Fig. \ref{fig:CoverChange}), and reduced the underestimates of cover (Fig. \ref{fig:IBM1step}). The version of the model with maximum removal treatment effects (removal coefficients set to the 95\% confidence limit furthest from zero), were qualitatively similar, but further amplified the contrast between predictions of the baseline and removal models for and \textit{P. spicata} REF MAX CI FIG

@@

% add max CI figure in SI

To quantify the long-term population-level consequences of removals, we used an IPM to project stochastic equilibrium cover. We concentrated on grass species' responses to \textit{A. tripartita} removal, since the growth regressions indicated our baseline model might underestimate competitive release in this situation (our baseline vital rate models worked well in predicting \textit{A. tripartita} responses to grass removal). Using the baseline model, with no removal treatment coeffiicients, the IPM predicts little effect of \textit{A. tripartita} removal on the cover of any grass species (compare gray and brown boxes in Fig. \ref{fig:IPMresults}). Using a model with removal treatment coefficients, the IPM predicts little effect of shrub removal on \textit{Poa secunda} and \textit{H. comata}, but a 2-fold increase in the cover of \textit{P. spicata} (Fig. \ref{fig:IPMresults}). Running a version of the model with maximum removal treatment effects (removal coefficients set to the 95\% confidence limit furthest from zero), resulted in the same qualitative outcome, but a larger increase in \textit{P. spicata} % (Fig. \ref{fig:IPMresults-maxCI})

% add max CI figure in SI

@@

For three of the four species, we find little or no evidence that models based on observational data underestimate the intensity of interspecific competition. Removal treatment effects were never significant for \textit{H. comata}; our baseline model, which accounts for local neighborhood composition, adequately explained vital rate responses to \textit{A. tripartita} removal. The same was true for the survival and growth models for \textit{A. tripartita}, but at the recruitment stage the grass removal treatment effect was significant and negative, implying that our baseline model might overestimate--not underestimate--competitive effects of the perennial grasses on \textit{Artemisia} recruitment. \textit{Poa secunda} showed mixed responses to \textit{Artemisia} removal: removal effects were not significant at the survival stage, were positive at the growth stage, and negative at the recruitment stage (where the estimated effect of \textit{Artemisia} on \textit{Poa} is positive ). At the population level, models with and without removal effects projected similar patterns of short (Fig. \ref{fig:CoverChange}) and long-term (Fig. \ref{fig:IPMresults}) dynamics for \textit{H. comata} and \textit{A. tripartita}. For \textit{Poa secunda}, the model with removal effects predicted higher population growth over the short but not the long-term, perhaps because the short-term projections mostly reflect the positive effect of \textit{Artemisia} removal on growth, while the long-term equilibrium are also sensitive to the negative effects of removal on recruitment. Although the baseline model did not always successfully predict removal effects on vital rates, it does appear to successfully capture the population-level consequences of removals on these three species and did not severly underestimate the magnitude of competitive release.

The story is different for the fourth species, \textit{P. spicata}. Both individual growth and short- and long-term population growth were higher in the sagebrush removal treatments than in the control plots, suggesting that our analyses of observational data may underestimate the competitive release that \textit{P. spicata} experiences following \textit{Artemisia} removal. However, the mechanism by which \textit{Artemisia} removal promotes \textit{P. spicata} remains unclear. If  \textit{Artemisia} removal is releasing \textit{P. spicata} from competition, we should expect \textit{P. spicata} individuals that were located beneath \textit{Artemisia} shrubs that we removed to respond more than individuals located away from any shrubs. Yet we found no evidence for such a difference. An alternative explanation involves pre-treatment differences in composition between control plots and  the \textit{Artemisia} removal  plots, which had higher cover of \textit{Artemisia} (see 2011 means in Fig. \ref{fig:CoverTrends}) and higher cover of subdominant grasses and shrubs (Fig. \ref{fig:W-by-treatment}). Perhaps these pre-treatment differences reflect subtle differences in edaphic and environmental factors which promote \textit{P. spicata} growth. While \textit{P. spicata} appears to show stronger competitive release than our baseline model predicted, the evidence is not conclusive.

The small competitive release of many species in our system implies that classic resource competition 
is not the primary factor determining community structure. 
\comment{SPE: I think this depends on what you mean by ``classic resource competition''. If you mean Tilman-style competition for one limiting resource distributed homogeneously, then I agree. But a MacArthur-Levins model with resource spectrum and distinct utilization curves could also be called classic resource competition,
and that model will have weak to nonexistent competitive release if niche overlap is low.} 
Interactions with above- or belowground enemies can also regulate populations and, given enough species-specificity in the effects of these enemies (Clark paper, Janzen, Connell), decouple interspecific dynamics. Species-specific responses to spatial and temporal environmental variation can also weaken interspecific interactions (Chesson et al. 2000), though in a previous study we failed to find evidence of a temporal storage effect in this community (Adler et al. 2009). More likely, plants may alter the environment, and even resource availability, in ways that facilitate their neighbors. For example, shade cast by \textit{Artemisia} canopies may reduce evapotranspiration (e.g. refs...). \textit{Artemisia} removal might offest this facilitative effect by reducing resource uptake, resulting in little net change in water balance following removal.  Unfortunately, we did not monitor soil moisture at the spatial and temporal resolution necessary to test this hypothesis.

Evidence for weak interspecific interactions and limited competitive release carries interesting implications. If most interspecific interactions are weak, then single-species models may be adequate for forecasting population responses to environmental change (Adler et al. 2012, Kleinhesselink and Adler 2015). Similarly, differences among species in relative abundance must reflect variation in the strength of intraspecific density dependence rather than variation in the strength of interspecific interactions (Yenni et al. 2015). In such a system, effective management and conservation requires understanding the sources and species-specificity of intraspecific density dependence. In fact, an experiment to test our model's skill at predicting release from intraspecific competition may be even more important than the test of interspecific competitive release described here.

Our results have given us more confidence that competitive release is quite small for some species in some communities. But for many species in other communities, interspecific competition may still be an important driver. In closed-canopy forests where competition for light is intense, we might have found different results than in our system, in which competition for a number of belowground resources is more important than competition for light. However, a number of recent studies from forests have found that heterospecific competitive effects are dramatically weaker than conspecific effects (Uriarte refs, Comita et al. 2010, Johnson et al. 2012). We also expect stronger competitive release in successional systems, or communities in the early stages of community assembly where the process of competitive exclusion is still underway (examples?). Finally, we might have found stronger competitive release if we could have studied rare species in our system, as long as those species are rare because they are competitively suppressed. If they are rare because they occupy unique, but narrow, niches, interspecific interactions would still be weak. Determining where interspecific interactions and competitive release are strong, and where they are not, remains a fascinating research problem. \comment{SPE: Equally interesting, I think, is the question of why interspecific
interactions are so weak (when they are), as opposed to what you would expect from ``all plant species compete for water, light, and a few mineral resources''.}

@@

>>>>

To quantify the short-term population-level consequences of removals, we used an IBM to project the cover of each species in each quadrat one year ahead using versions of the model with and without removal treatment effects. Our models tended to under-predict population growth (Fig. \ref{fig:CoverChange}) and population size (Fig. \ref{fig:IBM1step}) for all species, but the difference between predictions from the baseline and removal models varied among species. For \textit{A. tripartita}, both observed and predicted population growth rates were close to zero, and differences in model predictions were small. \textit{H. comata} experienced greater variation in growth rates, but the inclusion of removal treatment effects had little effect on population growth rates. For \textit{Poa secunda} and especially \textit{P. spicata}, the model with removal effects consistently predicted higher population growth than the baseline model (Fig. \ref{fig:CoverChange}), and reduced the underestimates of cover (Fig. \ref{fig:IBM1step}). The version of the model with maximum removal treatment effects (removal coefficients set to the 95\% confidence limit furthest from zero), were qualitatively similar, but further amplified the contrast between predictions of the baseline and removal models for and \textit{P. spicata} (Fig. \ref{fig:IBM1step-maxCI}).

@@

To quantify the long-term population-level consequences of removals, we used an IPM to project stochastic equilibrium cover. We concentrated on grass species' responses to \textit{A. tripartita} removal, since the growth regressions indicated our baseline model might underestimate competitive release in this situation (our baseline vital rate models worked well in predicting \textit{A. tripartita} responses to grass removal). Using the baseline model, with no removal treatment coeffiicients, the IPM predicts little effect of \textit{A. tripartita} removal on the cover of any grass species (compare gray and brown boxes in Fig. \ref{fig:IPMresults}). Using a model with removal treatment coefficients, the IPM predicts little effect of shrub removal on \textit{Poa secunda} and \textit{H. comata}, but a 2-fold increase in the cover of \textit{P. spicata} (Fig. \ref{fig:IPMresults}). Running a version of the model with maximum removal treatment effects (removal coefficients set to the 95\% confidence limit furthest from zero), resulted in the same qualitative outcome, but a larger increase in \textit{P. spicata} (Fig. \ref{fig:IPMresults-maxCI}).

@@

For three of the four species, we find little or no evidence that models based on observational data underestimate the intensity of interspecific competition. Removal treatment effects were never significant for \textit{A. tripartita} or \textit{H. comata}; our baseline model, which accounts for local neighborhood composition, adequately explained vital rate responses to  removal.  \textit{Poa secunda} showed mixed responses to \textit{Artemisia} removal: removal effects were not significant at the survival stage, were positive at the growth stage, and negative at the recruitment stage (where the estimated effect of \textit{Artemisia} on \textit{Poa} is positive). At the population level, models with and without removal effects projected similar patterns of short (Fig. \ref{fig:IBM1step}) and long-term (Fig. \ref{fig:IPMresults}) dynamics for \textit{H. comata} and \textit{A. tripartita}. For \textit{Poa secunda}, the model with removal effects predicted higher population growth over the short but not the long-term, perhaps because the short-term projections mostly reflect the positive effect of \textit{Artemisia} removal on growth, while the long-term equilibrium are also sensitive to the negative effects of removal on recruitment. Although the baseline model did not always successfully predict removal effects on vital rates, it does appear to successfully capture the population-level consequences of removals on these three species and did not severely underestimate the magnitude of competitive release.

The story is different for the fourth species, \textit{P. spicata}. Both individual growth and short- and long-term population growth were higher in the sagebrush removal treatments than in the control plots, suggesting that our analyses of observational data may underestimate the competitive release that \textit{P. spicata} experiences following \textit{Artemisia} removal. However, the mechanism by which \textit{Artemisia} removal promotes \textit{P. spicata} remains unclear. If  \textit{Artemisia} removal is releasing \textit{P. spicata} from competition, we should expect \textit{P. spicata} individuals that were located beneath \textit{Artemisia} shrubs that we removed to respond more than individuals located away from any shrubs. Yet we found no evidence for such a difference. An alternative explanation involves pre-treatment differences in composition between control plots and  the \textit{Artemisia} removal  plots, which had higher cover of \textit{Artemisia} (see 2011 means in Fig. \ref{fig:CoverTrends}) and higher cover of subdominant grasses and shrubs (Fig. \ref{fig:W-by-treatment}). Perhaps these pre-treatment differences reflect subtle differences in edaphic and environmental factors which could promote \textit{P. spicata} growth. While \textit{P. spicata} appears to show stronger competitive release than our baseline model predicted, the evidence is not conclusive.

The small competitive release of many species in our system implies that each of the four dominant species we studied are limited by quite different factors. This low niche overlap could reflect strong resource partitioning (Tilman 1982), perhaps combined with variation in the times and locations at which species are taking up those resources. Species-specific responses to spatial and temporal variation in non-resource environmental variables can also weaken interspecific interactions (Chesson et al. 2000), though in a previous study we failed to find evidence of a temporal storage effect in this community (Adler et al. 2009). Interactions with above- or belowground enemies can regulate populations and, given enough species-specificity in the effects of these enemies (Clark paper, Janzen, Connell), decouple interspecific dynamics.  More likely, plants may alter the environment, and even resource availability, in ways that facilitate their neighbors. For example, shade cast by \textit{Artemisia} canopies may reduce evapotranspiration (e.g. refs...). \textit{Artemisia} removal might offest this facilitative effect by reducing resource uptake, resulting in little net change in water balance following removal.  Unfortunately, we did not monitor soil moisture at the spatial and temporal resolution necessary to test this hypothesis.

Evidence for weak interspecific interactions and limited competitive release carries interesting implications. If most interspecific interactions are weak, then single-species models may be adequate for forecasting population responses to environmental change (Adler et al. 2012, Kleinhesselink and Adler 2015). Similarly, differences among species in relative abundance should reflect variation in the strength of intraspecific density dependence more than variation in the strength of interspecific interactions (Yenni et al. 2015). In such a system, effective management and conservation requires understanding the sources and species-specificity of intraspecific density dependence. In fact, an experiment to test our model's skill at predicting release from intraspecific competition could be more important than the test of interspecific competitive release described here.

Our results have given us more confidence that competitive release is quite small for some species in some communities. But for many species in other communities, interspecific competition may still be an important driver. In closed-canopy forests where competition for light is intense, we might have found different results than in our system, in which competition for a number of belowground resources is more important than competition for light. However, a number of recent studies from forests have found that heterospecific competitive effects are dramatically weaker than conspecific effects (Uriarte refs, Comita et al. 2010, Johnson et al. 2012). We also expect stronger competitive release in successional systems, or communities in the early stages of community assembly where the process of competitive exclusion is still underway (examples?). Finally, we might have found stronger competitive release if we could have studied rare species in our system, as long as those species are rare because they are competitively suppressed. If they are rare because they occupy unique, but narrow, niches, interspecific interactions would still be weak. Determining why interspecific interactions and competitive release are strong at some times and places and why they are weak in other situations remains a fascinating research problem.

@@

 \begin{figure}[tbp]
 \centering
 \includegraphics[width=1\textwidth]{cover_projections_1step_maxCI}
 \caption{Observed and predicted cover for each of the four modeled species assuming maximum effects of removal. Solid lines show mean observed cover averaged across control (blue) and removal treatment (red) quadrats. Dashed lines show one-step-ahead predictions from the baseline IBM (no removal treatment coefficients), averaged across quadrats. Dotted lines show one-step-ahead predictions from an IBM that includes removal treatment coefficients representing the outermost values of the 95\% confidence intervals. }
 \label{fig:IBM1step-maxCI}
 \end{figure}

  \begin{figure}[tbp]
  \centering
  \includegraphics[width=1\textwidth]{boxplots-maxCI}
  \caption{Equilibrium cover of the four dominant species simulated by the IPM with maximum removal treatment effects. Boxplots show interannual variation reflecting random year effects. Gray boxes show cover simulated by a model with \textit{A. tripartita} present and no removal treatment coefficients, brown boxes show results from the same model but with \textit{A. tripartita} cover set to zero (a species removal), green boxes show results from a model with  \textit{A. tripartita} set to zero and (mean) removal treatment coefficients included, and blue boxes show results with removal treatments coefficients assigned the outermost value of the 95\% confidence intervals.  }
  \label{fig:IPMresults-maxCI}
  \end{figure}
